The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-12 src/ms.tex
edits through section 3

% Much of this text was adapted from Robel Geda's report from NSBP and
% SACNAS. Please include him on the author list.
\textbf{\astropy representation at national diversity conferences:}
Several members of the \astropy community attended virtual conferences
of the National Society of Black Physicists (NSBP) and SACNAS
(focusing on Hispanic and Native American scientists across all STEM
fields). \astropy representatives noted that one underlying topic came
up multiple times. A major barrier for persons who come from
underrepresented communities is the lack of resources and expertise
that are necessary to train students. Writing PEP 8 compliant software
and understanding the Git and Github workflow is not part of the
standard Physics and Astrophysics curriculum. Additionally, the larger
Astronomnical community is still in the process of transitioning
towards Python based tools, making it difficult for students not
currently under advisement by some one with extensive Python expertise
to get involved with the open source Python community. Projects such
as Learn Astropy could have a profound impact by empowering
underrepresented groups because it provides a free, searchable, and
accessible introduction to Python tools for astrophysical
research. Representatives also noted that offering training and
teaching materials to PIs of Research Experience for Undergraduate
(REU) programs throughout the United States could be helpful for
encouraging advisors to teach students how to use Astropy, create
their own libraries, and use version control on open source platforms
like Github. Such materials could also be offered as a workshop at
national diversity conferences themselves, as SACNAS solicits special
session proposals each year.

\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
  initiative:} \todo{No author assigned}{Describe more about this.}

\subsection{Funding} \label{sec:project-funding}

\secauthor{Aarya Patil}

@@

With the goal of establishing a sustainable financial model for the Astropy
Project, the Coordination Committee under the new governance charter (see \S
\ref{sec:project-governance}) established a standing Finance committee.
This committee oversee the planning and allocation of finances on behalf
of the project. The Astropy Project accepts funds from institutions as well as
individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
supports and promotes world-class, innovative, open source scientific computing.}.
Previously, no direct financial support was available for the project development,
and NUMFOCUS covered most of the incurred operational costs. However, transitioning
to long-term financial stability meant having an influx of funds in the form of
institutional support, and hence the Astropy Project applied for and successfully
acquired four grants. These grants are administered by NumFOCUS on behalf of the
project, providing a ``neutral'' space not tied to any particular astronomy
institution. The two major ones are briefly summarized below.

\begin{itemize}
    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project". Over a three-year term, this grant supported
    sustenance of the project by providing funds for (1) the development and
    maintenance of the \astropypkg and its infrastructure to existing project
    members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the formalization
    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
    improvement of equity, diversity, and inclusion efforts in the project (see \S
    \ref{sec:project-ide}). Some of the money supported travel for meetings,
    conferences, and workshops. This generous grant helped reduce reliance on
    volunteer-driven maintenance of the openly-developed package and paved
    the way to acquire funding from federal agencies: a big milestone for the
    success of the project.

    \item NASA: A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages that need support, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{itemize}

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All these successful proposals show that the Astropy project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

>>>>

% Much of this text was adapted from Robel Geda's report from NSBP and
% SACNAS. Please include him on the author list.
\textbf{\astropy representation at national diversity conferences:}
Several members of the \astropy community attended virtual conferences of the
National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
Native American scientists across all STEM fields). \astropy representatives
noted that one underlying topic came up multiple times. A major barrier for
persons who come from underrepresented communities is the lack of resources and
expertise that are necessary to train students. Writing PEP 8 compliant software
and understanding the Git and \github workflow is not part of the standard
Physics and Astrophysics curriculum. Additionally, the larger astronomical
community is still in the process of transitioning towards \python-based tools,
making it difficult for students not currently under advisement by some one with
extensive \python expertise to get involved with the open source \python
community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
have a profound impact by empowering underrepresented groups because it provides
a free, searchable, and accessible introduction to \python tools for
astrophysical research. Representatives also noted that offering training and
teaching materials to PIs of Research Experience for Undergraduate (REU)
programs throughout the United States could be helpful for encouraging advisors
to teach students how to use Astropy, create their own libraries, and use
version control on open source platforms like \github. Such materials could also
be offered as a workshop at national diversity conferences themselves, as SACNAS
solicits special session proposals each year.

% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
%   initiative:} \todo{No author assigned}{Describe more about this.}

\subsection{Project funding} \label{sec:project-funding}

% \secauthor{Aarya Patil}

@@

With the goal of establishing a sustainable financial model for the Astropy
Project, the Coordination Committee, under the new governance charter (see
Section~\ref{sec:project-governance}), established a standing Finance Committee.
This committee oversees the planning and allocation of finances on behalf
of the Project. The Astropy Project accepts funds from institutions as well as
individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
supports and promotes world-class, innovative, open source scientific computing.}
Previously, no direct financial support was available for the project development,
and NumFOCUS covered most of the incurred operational costs. However, transitioning
to long-term financial stability meant having an influx of funds in the form of
institutional support, and hence the Astropy Project applied for and successfully
acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
Project, providing a ``neutral'' space not tied to any particular astronomy
institution. The two major grants are briefly summarized below.

\begin{description}
    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project.'' Over a three-year term, this grant supported
    sustenance of the Project by providing funds for (1) the development and
    maintenance of the \astropypkg and its infrastructure, provided to existing
    project members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the
    formalization of a governance structure (see
    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
    diversity, and inclusion efforts in the project (see
    Section~\ref{sec:project-ide}). Some of this money supported travel for
    meetings, conferences, and workshops. This generous grant helped reduce
    reliance on volunteer-driven maintenance of the openly-developed package and
    paved the way to acquire funding from federal agencies, representing a major
    milestone for the success of the project.

    \item[NASA] A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{description}

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All of these successful proposals show that the Astropy Project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.
